Short Paper

### 2023-04-05 manuscript/manuscript.qmd
Figures and tables added

---
title: Short Paper
subtitle: A Short Subtitle
author:
  - name: Alice Anonymous
    email: alice@example.com
    affiliations: 
        - id: some-tech
          name: Some Institute of Technology
          department: Department Name
          address: Street Address
          city: City
          state: State
          postal-code: Postal Code
    attributes:
        corresponding: true
    note: This is the first author footnote.
  - name: Bob Security
    email: bob@example.com
    affiliations:
        - id: another-u
          name: Another University
          department: Department Name
          address: Street Address
          city: City
          state: State
          postal-code: Postal Code
    note: |
      Another author footnote, this is a very long footnote and it should be a really long footnote. But this footnote is not yet sufficiently long enough to make two lines of footnote text.
  - name: Cat Memes
    email: cat@example.com
    affiliations:
        - ref: another-u
    note: Yet another author footnote.
  - name: Derek Zoolander
    email: derek@example.com
    affilations:
        - ref: some-tech
abstract: |
  This is the abstract. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Vestibulum augue turpis, dictum non malesuada a, volutpat eget velit. Nam placerat turpis purus, eu tristique ex tincidunt et. Mauris sed augue eget turpis ultrices tincidunt. Sed et mi in leo porta egestas. Aliquam non laoreet velit. Nunc quis ex vitae eros aliquet auctor nec ac libero. Duis laoreet sapien eu mi luctus, in bibendum leo molestie. Sed hendrerit diam diam, ac dapibus nisl volutpat vitae. Aliquam bibendum varius libero, eu efficitur justo rutrum at. Sed at tempus elit.
keywords: 
  - keyword1
  - keyword2
date: last-modified
bibliography: man-qechairquality.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Habitat International
      formatting: review
      model: 3p
      cite-style: authoryear
      highlights:
        - Highlight 1
        - Highlight 2
        - Highlight 3
editor: 
  markdown: 
    wrap: 72
---

@@

Queen's, simply, produces an immense amount of waste that it is unable
to efficiently manage. The dozens of wards, offices, and kitchens,
serving the hundreds of patients and staff, generates considerable
infectious medical waste, mixed domestic and office waste, and kitchen
waste, which needs to be removed and disposed of daily. Some individual
wards implement waste separation, binning medical wastes separately from
general waste, however separation ends at the point of collection.
Queen's is not served by regular municipal waste collection. What
municipal collection that is done is ad-hoc, sporadic, and based on the
hospital's available financial resources. Rather, most hospital waste
(infectious and non-infectious) is gathered by grounds staff and burnt
at the hospital's incinerator, located at a central point within the
hospital campus[^1] [@kalina2020this]. However, the vast majority of the
deposited waste is not incinerated, as a new incinerator provided by the
Ministry of Health in late 2019, is only able to handle a small
percentage of the hospital's waste. Furthermore, the incinerator is
frequently inoperable, under repair, or without power[^2]. Rather, the
waste is openly burned, in a constantly smouldering pile to the side of
the incinerator building, sending up the aforementioned clouds of smoke,
which choke the grounds (Figure 1). Innumerable other, smaller, fires,
scattered across the QECH campus (Figure 2) contribute their part, as
grounds and maintenance staff habitually burn leaves and other garden
refuse. Furthermore, caregivers[^3] who reside separately on hospital
grounds, cook for themselves throughout the day and night using biomass
(charcoal or wood), and burn their own domestic waste. As a result, air
quality at the hospital is a constant source of discussion, with
patients, staff, and caregivers struggling to cope, when just breathing
is, as the quote used in the title of this article, 'unbearable.'

@@

```{r}
# Figure 1: Open burning at QECH, with the old incinerator building in the background (Authors).
knitr::include_graphics(here::here("manuscript/img/image-replace-tree.png"))
```

@@

```{r}
# map
# Figure 2: Sensor locations within QECH
```

@@

Therefore, for international comparisons and due to the lack of
comprehensive Malawian Standards, we use the WHO recommendations as a
basis of comparison for the measured values
[@worldhealthorganization2021who]. The relevant particulate matter
targets are presented in Table 1.

```{r}
# Table 1. Particulate matter targets from @worldhealthorganization2021who
knitr::include_graphics(here::here("manuscript/img/image-replace-table.png"))

@@

Due to localised variation (air movement in the immediate area) and the
density of the observations, the plotted 5-minute data obscure
persistent trends. The full set of 5-minute data are plotted and
presented in Appendix A Figure A.1 (TODO). However, the number of measured values
that exceeded the hazardous limit for both parameters are summarised in
Table 2.

```{r}
# Table 2. Peaks for both PM2.5 and PM10 are the number of data points above the WHO interim target 1 (annual) which is the least stringent;  the number of observations recorded (count) is provided for reference.

@@

The 24-hour averaged values (logged every 5 minutes) are presented for
each location, for both PM10 and PM2.5 in Figure 3. As well as dampening
local variation and peaks, the WHO air quality guidelines are also based
on 24-hour averages which are the standard against which the health
risks can be judged.

@@

```{r}
# Figure 3. Average 24-hour PM10 and PM2.5 at 8 monitoring stations over 2 months

@@

Given that the type of fuel and the type of burning (contained vs. open)
produce very different particulate "fingerprints", the ratio of PM10 to
PM2.5 values were examined to determine if a clear difference between
locations, and therefore source, could be identified. The results are
presented in Figure 4.

```{r}
# Figure 4. PM10:PM2.5 values for each location.  Each line shows the minimum and maximum ratio for the day based on ratios calculated hourly.  
```

@@

The relative amount of time exposed to a given category of air quality
is shown in Figure 5. Each square represents 1% of the measured time.

```{r}
# Figure 5. Percentage of measured values categorised according to WHO 2021 targets
```

@@

```{r}

# Figure XA1. PM2.5 and PM10 values collected every 5 minutes over 10 months at 8 locations. 
# TODO: ##Lars can you please fix the x-axis labels so they aren’t squished

@@

# Figures and tables

@fig-meaningless is generated using an R chunk.

```{r}
#| label: fig-meaningless
#| fig-cap: A meaningless scatterplot
#| fig-width: 5
#| fig-height: 5
#| fig-align: center
#| out-width: 50%
#| echo: false
plot(runif(25), runif(25))
```

# Tables coming from R

Tables can also be generated using R chunks, as shown in @tbl-simple
example.

```{r}
#| label: tbl-simple
#| tbl-cap: Caption centered above table
#| echo: true
knitr::kable(head(mtcars)[,1:4])
```

>>>>

---
title: Short Paper
subtitle: A Short Subtitle
author:
  - name: Alice Anonymous
    email: alice@example.com
    affiliations: 
        - id: some-tech
          name: Some Institute of Technology
          department: Department Name
          address: Street Address
          city: City
          state: State
          postal-code: Postal Code
    attributes:
        corresponding: true
    note: This is the first author footnote.
  - name: Bob Security
    email: bob@example.com
    affiliations:
        - id: another-u
          name: Another University
          department: Department Name
          address: Street Address
          city: City
          state: State
          postal-code: Postal Code
    note: |
      Another author footnote, this is a very long footnote and it should be a really long footnote. But this footnote is not yet sufficiently long enough to make two lines of footnote text.
  - name: Cat Memes
    email: cat@example.com
    affiliations:
        - ref: another-u
    note: Yet another author footnote.
  - name: Derek Zoolander
    email: derek@example.com
    affilations:
        - ref: some-tech
abstract: |
  This is the abstract. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Vestibulum augue turpis, dictum non malesuada a, volutpat eget velit. Nam placerat turpis purus, eu tristique ex tincidunt et. Mauris sed augue eget turpis ultrices tincidunt. Sed et mi in leo porta egestas. Aliquam non laoreet velit. Nunc quis ex vitae eros aliquet auctor nec ac libero. Duis laoreet sapien eu mi luctus, in bibendum leo molestie. Sed hendrerit diam diam, ac dapibus nisl volutpat vitae. Aliquam bibendum varius libero, eu efficitur justo rutrum at. Sed at tempus elit.
keywords: 
  - keyword1
  - keyword2
date: last-modified
bibliography: man-qechairquality.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Habitat International
      formatting: review
      model: 3p
      cite-style: authoryear
      highlights:
        - Highlight 1
        - Highlight 2
        - Highlight 3
execute: 
  echo: false
  warning: false
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: console
---

```{r setup}

# Source data preparation script
source(here::here("R/01-data-preparation.R"))

@@

Queen's, simply, produces an immense amount of waste that it is unable
to efficiently manage. The dozens of wards, offices, and kitchens,
serving the hundreds of patients and staff, generates considerable
infectious medical waste, mixed domestic and office waste, and kitchen
waste, which needs to be removed and disposed of daily. Some individual
wards implement waste separation, binning medical wastes separately from
general waste, however separation ends at the point of collection.
Queen's is not served by regular municipal waste collection. What
municipal collection that is done is ad-hoc, sporadic, and based on the
hospital's available financial resources. Rather, most hospital waste
(infectious and non-infectious) is gathered by grounds staff and burnt
at the hospital's incinerator, located at a central point within the
hospital campus[^1] [@kalina2020this]. However, the vast majority of the
deposited waste is not incinerated, as a new incinerator provided by the
Ministry of Health in late 2019, is only able to handle a small
percentage of the hospital's waste. Furthermore, the incinerator is
frequently inoperable, under repair, or without power[^2]. Rather, the
waste is openly burned, in a constantly smouldering pile to the side of
the incinerator building, sending up the aforementioned clouds of smoke,
which choke the grounds (@fig-open-burning). Innumerable other, smaller, fires,
scattered across the QECH campus (@fig-map-sensors) contribute their part, as
grounds and maintenance staff habitually burn leaves and other garden
refuse. Furthermore, caregivers[^3] who reside separately on hospital
grounds, cook for themselves throughout the day and night using biomass
(charcoal or wood), and burn their own domestic waste. As a result, air
quality at the hospital is a constant source of discussion, with
patients, staff, and caregivers struggling to cope, when just breathing
is, as the quote used in the title of this article, 'unbearable.'

@@

```{r}
#| label: fig-open-burning
#| fig-cap: Open burning at QECH, with the old incinerator building in the background (Authors).
#| fig-align: center
#| out-width: 80%

knitr::include_graphics(here::here("manuscript/img/open-waste-burning-incinerator-qech.jpg"))
```

@@

```{r}
#| label: prepare-sensor-map

# Icon from: https://github.com/apancik/public-domain-icons/blob/master/dist/symbol%20fire.svg

fire_icon <- makeIcon(iconUrl = here::here("manuscript/img/symbol-fire.png"),
                      iconWidth = 50,
                      iconHeight = 50)

# Create a palette that maps factor levels to colors
pal <- colorFactor(c("#EE4B2B", "navy"), domain = c("sensor", "fire"))

fire_loc <- locations  |>
  filter(type == "fire")

sensor_loc <- locations  |>
  filter(type == "sensor")

qech_sensor_map <- leaflet(sensor_loc)  |>
  setView(lng = 35.021071999805755,
          lat = -15.803269709021687,
          zoom = 18)  |>
  #addProviderTiles(providers$Stamen)  |>
  addTiles()  |>
  addMarkers(
    lng = fire_loc$long,
    lat = fire_loc$lat,
    label = fire_loc$location,
    icon = ~fire_icon,
    labelOptions = labelOptions(noHide = T,
                                textsize = "15px",
                                direction = "bottom",
                                offset = c(0, 20),
                                style = list(
                                  "box-shadow" = "3px 3px rgba(0,0,0,0.25)",
                                  "border-color" = "rgba(0,0,0,0.5)"
                                  #"font-style" = "italic"
                                ))
  )  |>
  addCircleMarkers(
    lng = ~long,
    lat = ~lat,
    label = ~location,
    radius = 16,
    stroke = FALSE,
    fillOpacity = 1,
    color = ~pal(type),
    #icon = ~fire_icon,
    labelOptions = labelOptions(noHide = T,
                                textsize = "15px",
                                direction = "bottom",
                                offset = c(0, 10),
                                style = list(
                                  "box-shadow" = "3px 3px rgba(0,0,0,0.25)",
                                  "border-color" = "rgba(0,0,0,0.5)"
                                  #"font-style" = "italic"
                                )))

qech_sensor_map |> 
  mapshot(file = here::here("manuscript/figs/map-sensors-qech-blantyre.png"))
```

```{r}
#| label: fig-map-sensors
#| fig-cap: Sensor locations within QECH
#| fig-align: center
#| out-width: 80%

# Figure 2: Sensor locations within QECH
knitr::include_graphics(here::here("manuscript/figs/map-sensores-qech-blantyre.png"))
```

@@

Therefore, for international comparisons and due to the lack of
comprehensive Malawian Standards, we use the WHO recommendations as a
basis of comparison for the measured values
[@worldhealthorganization2021who]. The relevant particulate matter
targets are presented in @tbl-who-targets.

 (TODO: REPLACE screenshot)

```{r}
#| label: tbl-who-targets
#| tbl-cap: Particulate matter targets from [@worldhealthorganization2021who]

knitr::include_graphics(here::here("manuscript/img/image-replace-table.png"))

@@

Due to localised variation (air movement in the immediate area) and the
density of the observations, the plotted 5-minute data obscure
persistent trends. The full set of 5-minute data are plotted and
presented in Appendix A Figure A.1. However, the number of measured values
that exceeded the hazardous limit for both parameters are summarised in
@tbl-peaks.

```{r}
#| label: tbl-peaks
#| tbl-cap: Peaks for both PM~2.5~ and PM~10~ are the number of data points above the WHO interim target 1 (annual) which is the least stringent; the number of observations recorded (count) is provided for reference.

observations_location <- dat_in |> 
  select(date_time, location) |> 
  unique() |> 
  count(location)

no_peaks_who <- dat_in |> 
  mutate(peaks = case_when(
    indicator == "pm2.5" & value >  35 ~ TRUE,
    indicator == "pm10" & value >  70 ~ TRUE,
    TRUE ~ FALSE
  )) |> 
  mutate(date = as_date(date_time)) |> 
  group_by(indicator, location) |> 
  summarise(
    peaks = sum(peaks)
  ) |> 
  ungroup() |> 
  pivot_wider(names_from = indicator, values_from = peaks)

observations_location |> 
  left_join(no_peaks_who) |> 
  knitr::kable()
```

@@

The 24-hour averaged values (logged every 5 minutes) are presented for
each location, for both PM10 and PM2.5 in fig-24-hour-average. As well as dampening
local variation and peaks, the WHO air quality guidelines are also based
on 24-hour averages which are the standard against which the health
risks can be judged.

@@

```{r}
#| label: fig-24-hour-average
#| fig-cap: Average 24-hour PM10 and PM2.5 at 8 monitoring stations over 2 months
#| fig-align: center

dat_in_sum_day |> 
  ggplot(aes(x = date, y = mean, group = indicator, color = indicator)) +
  geom_point(size = 0.8) +
  geom_line() +
  labs(x = NULL) +
  ylab(bquote(uq/m^3)) +
  facet_wrap(~location, ncol = 4) + 
  scale_y_continuous(limits = c(0, 400)) +
  scale_color_brewer(type = "qual", palette = 2) + 
  theme_bw(base_size = 14) + 
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank())

@@

Given that the type of fuel and the type of burning (contained vs. open)
produce very different particulate "fingerprints", the ratio of PM10 to
PM2.5 values were examined to determine if a clear difference between
locations, and therefore source, could be identified. The results are
presented in @fig-pm-ratios.

```{r}
#| label: fig-pm-ratios
#| fig-cap: PM10:PM2.5 values for each location. Each line shows the minimum and maximum ratio for the day based on ratios calculated hourly. Does not include 14th to 16th October at location uMoyo due to extreme outliers.
#| fig-align: center

dat_in_ratios <- dat_in |> 
  mutate(date = as_date(date_time)) |> 
  mutate(hour = hour(date_time)) |> 
  group_by(date, hour, location, indicator) |> 
  summarise(mean = mean(value)) |> 
  pivot_wider(names_from = indicator, 
              values_from = mean) |> 
  mutate(p10_pm2.5_ratio = pm10/pm2.5)  |>

  filter((date != "2019-10-14" & date != "2019-10-15" & date != "2019-10-16")
         | location != "uMoyo") |> 
  #filter(p10_pm2.5_ratio > 30)

  group_by(date, location) |> 
  summarise(min = min(p10_pm2.5_ratio),
            max = max(p10_pm2.5_ratio))

# Plot

dat_in_ratios |> 
  ggplot(aes(x = date)) +
  geom_linerange(aes(ymin = min, ymax = max)) +
  labs(x = "hour") +
  facet_wrap(~location, ncol = 2)
```

@@

The relative amount of time exposed to a given category of air quality
is shown in @fig-percent-exposure. Each square represents 1% of the measured time.

```{r}
#| label: fig-percent-exposure
#| fig-cap: Percentage of measured values categorised according to WHO 2021 targets
#| fig-align: center

dat_air_quality_sum_who_2021 |> 
  #filter(indicator == "pm2.5") |> 
  group_by(location, indicator) |> 
  mutate(exposure_percent = round(exposure_hrs / sum(exposure_hrs) * 100, 0)) |>

  # manipulate data for purpose of plot

  mutate(exposure_percent = case_when(
    location == "Administration" 
    & indicator == "pm10"
    & air_quality_who_annual == "Interim target 2" ~ exposure_percent - 1,

    location == "Administration" 
    & indicator == "pm2.5"
    & air_quality_who_annual == "Interim target 3" ~ exposure_percent + 1,

    location == "uMoyo" 
    & indicator == "pm2.5"
    & air_quality_who_annual == "Interim target 4" ~ exposure_percent - 1,

    location == "Mercy James" 
    & indicator == "pm10"
    & air_quality_who_annual == "Interim target 4" ~ exposure_percent + 1,

    TRUE ~ exposure_percent
  )) |>

  #filter(is.na(exposure_percent))

ggplot(aes(fill = air_quality_who_annual, values = exposure_percent)) +
  geom_waffle(color ="white", size = 0.5, n_rows = 5, na.rm = TRUE) +
  coord_equal() +
  labs(fill = "WHO 2021 targets (annual)") +
  scale_fill_brewer(type = "div", palette = "RdYlBu", direction = -1) +
  scale_x_discrete(expand=c(0,0)) +
  scale_y_discrete(expand=c(0,0)) +
  facet_grid(location ~ indicator, switch = "y") +

  #facet_wrap(~location, ncol = 2) +
  theme_minimal(base_size = 14) +
  # help from: https://stackoverflow.com/questions/34749636/left-justify-text-from-multi-line-facet-labels
  theme(strip.text.x = element_text(hjust = 0),
        strip.text.y = element_text(vjust = 0, size = 4))
```

@@

```{r}
#| label: fig-5-minute-intervals
#| fig-cap: PM2.5 and PM10 values collected every 5 minutes over 10 months at 8 locations.
#| fig-align: center

dat_in |> 
  ggplot(aes(x = date_time, y = value)) +
  geom_line(alpha = 0.5) +
  labs(title = "Hospital Air Quality",
       subtitle = "PM2.5 and PM10 measurements in 5-min intervals at eight locations") +
  ylab(bquote(uq/m^3)) +
  facet_grid(indicator~location) +
  theme_bw(base_size = 14) + 
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

@@
